The Functional Therapeutic Chemical Classification System

### 2013-10-09 document.tex
stop reviews

\subsubsection{Polypharmacology spectrum}
The more information on a drug’s molecular targets and their physiological roles, the more opportunities 
exist to re-orient a drug into doing something new. The therapeutic agents described in the FTC can have several MoAs, 
representing the intrinsic polypharmacology of the approved compounds. Figure \ref{fig:03} illustrates the polypharmacology 
spectrum by showing the distribution of number of MoAs per compound. When only direct categories are considered, compounds belong on average
to 13.5 MoA categories. This number increases to 61.2 when parent categories are taken into consideration (super classes).
Not all the MoAs are disease relevant, 
some FTC categories are particularly abstract (e.g, \emph{Anti-biological process agent}) yet they represent discrete 
categories to which the drug belongs with an explicit and clear meaning. These discrete MoAs are a good starting 
point to understand what a compound can do when administered in a human system, as it will be shown with Alzheimer's disease.
The human readable definitions 
associated to each FTC category furthermore help the user to choose the adequate MoA for the desired outcome.
Compound's polypharmacology is 
well represented in the FTC, as shown by the numerous MoAs each approved drug can exhibit.

@@

The FTC is a novel classification for approved drugs, which can be used as a starting point to 
generate drug repurposing hypotheses. We have guided the reader through some of the ways the resource 
can be explored. No experimental validations have been presented in this manuscript, yet references to 
supporting literature was provided in order to legitimate the claims. The main goal of this document is to 
introduce the resource to the community; deeper analyses are being performed over the classification at the time of writing.

>>>>

\subsubsection{Polypharmacology spectrum}
The more information on a drug’s molecular targets and their physiological roles, the more opportunities 
exist to re-orient a drug into doing something new. The therapeutic agents described in the FTC can have several MoAs, 
representing the intrinsic polypharmacology of the approved compounds. Figure \ref{fig:03} illustrates the polypharmacology 
spectrum by showing the distribution of number of MoAs per compound. When only direct categories are considered, compounds belong on average
to 13.5 MoA categories. This number increases to 61.2 when parent categories are taken into consideration (super classes).
Not all the MoAs are disease relevant, 
some FTC categories are particularly abstract (e.g, \emph{Anti-biological process agent}) yet they represent discrete 
categories to which the drug belongs with an explicit and clear meaning. These discrete MoAs are a good starting 
point to understand what a compound can do when administered in a human system.
Compound's polypharmacology is 
well represented in the FTC, as shown by the numerous MoAs each approved drug can exhibit.
We decided to further look at a well-known repositioning example, 
in order to see whether the FTC was suitable to identify the new
uses of an old drug. We picked the \emph{thalidomide} for 
this exercise ({{https://www.ebi.ac.uk/chembl/ftc/agent/DB01041}}). The molecule was 
first indicated to treat morning sickness
in pregnant women, but got quickly withdrawn from the market after the dramatic identification 
of teratogenic effect in newborns, caused by the drug. The accepted molecular mechanism behind the side-effect is  
an impairment of the angiogenic process responsible for the development of members, affecting in particular the limbs \citep{therapontos2009thalidomide}.
The \emph{thalidomide} was accurately classified as \emph{Anti-cell migration involved in sprouting angiogenesis agent} in the FTC, capturing the known toxicity of the drug. The \emph{thalidomide} is currently investigated for a multitude of new usages,
in particular for anti-cancer and immunomodulatory activities among others \citep{teo2005thalidomide}.
These new indications are well represented in the FTC too, for 
example by the categories \emph{Anti-vascular endothelial growth factor production agent} or 
\emph{Anti-cell division agent} for neoplastic activities, 
or by the classes \emph{Anti-cytokine secretion agent} and \emph{Anti-I-kappaB kinase/NF-kappaB cascade agent} for its effect
on the immune system. These observations demonstrate that the FTC can successfully capture the 
molecular reasons behind the repositioning of an old compound. 
Moreover the classification can also provide valuable insight regarding
potential toxicity too.

@@

The FTC is a novel classification for approved drugs, which can be used as a starting point to 
generate drug repurposing hypotheses. We have guided the reader through some of the ways the resource 
can be explored. No experimental validations have been presented in this manuscript, yet references to 
supporting literature was provided in order to legitimate the claims. The main goal of this document is to 
introduce the resource to the community; deeper analyses are being performed over the classification at the time of writing.
This classification leverages the information present in various databases and ontologies, similarly to the 
Open PHACTS initiative \citep{williams2012open} and to the work done by \cite{hoehndorf2012identifying}. The FTC
mostly differentiates itself from these projects by providing a whole new set of novel categories, 
dedicated to tackle a very specific problem: drug repositioning.
